---
title: "Chapter 5: Data Transformation"
output: 
  html_document: 
    keep_md: yes
---
```{r, include=FALSE}
library(nycflights13)
library(tidyverse)
```
In-class notes

`x %in% y` code will select every row where x is a value in y.

Missing values: confusing. NA is an unknown number. NaN is a nonsense result like 1/0 whose answer is not a real number.

Use `$` to indicate wanting one column like `flights$dep_time` but also in `tidyverse` you don't have to do this because you already called flights in these functions.

`lead()` and `lag()` return the vector offset by 1 either at the front or the end and adds NA there.

`%>%` is a pipe function. So in the terminal it's just `>`or `|`.

For changing the order of the columns, can use `arrange()` but also `mutate()`.


Note: Variable Types  
`int` = integers  
`dbl` = real numbers  
`chr` = strings  
`dttm` = date-time  
`lgl` = logical
`fctr` = categorical variables with fixed possible values  
`date` = dates  

Note: dplyr functions  
*Pick observations by value with `filter()`.
*Reorder rows with `select()`.
*Create new variables with functions of existing variables with `mutate()`.
*Collapse many values into a single summary with `summarise()`.
*Change the scope of the above functions from operating on entire dataset to group-by-group with `group()`.

These funtions all work similarly.  
1. first argument is a data frame  
2. subsequent arguments describe what to do with the data frame using variable names without quotes  
3. result is a new data frame  

### 5.2.4 Exercises  
**1. Find all the flights that**  
> 1. Had an arrival delay of two or more hours  
> 2. Flew to Houston (`IAH` or `HOU`)  
> 3. Were operated by United, American, or Delta  
> 4. Departed in summer (July, August, and September)  
> 5. Arrived more than two hours late, but didn't leave late  
> 6. Were delayed by at least an hour, but made up over 30 minutes in flight  
> 7. Departed between midnight and 6am (inclusive)

```{r}
filter(flights, arr_delay >= 120)
#an arrival delay of two or more hours, in minutes
```

```{r}
filter(flights, dest=='IAH' | dest=='HOU')
#destination of IAH OR HOU
```

```{r}
filter(flights, carrier=='UA' | carrier=='AA' | carrier=='DL')
#carrier United OR American OR Delta
```

```{r}
filter(flights, month %in% c(7,8,9))
#flights in months 7, 8, 9 (July, August, September)
```

```{r}
filter(flights, arr_delay >= 120 & dep_delay <= 0)
#flights that weren't delayed taking off but landed late
```

```{r}
filter(flights, dep_delay >= 60 & dep_delay- arr_delay > 30)
#flights that left more than an hour late but made up delay to within 30 minutes during flight
```

```{r}
filter(flights, dep_time == 2400 | dep_time <= 0600)
#flights that left at midnight OR before/at 6am in 24hr time
```


**2. Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?**
```{r}
filter(flights, between(dep_delay, 0, 120))
```

_This produces a tibble of flights that have a take off delay between 0 and 120 minutes._

**3. How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?**
Running `summary` on the data set will tell you how many NAs there are in `dep_time`.

```{r}
summary(flights)
```

_There are 8,255_ `NA` _in_ `dep_time`.

**4. Why is `NA^0` not missing? Why is `NA | TRUE` not missing? Why is `FALSE & NA` not missing? Can you figure out the general rule?**
_NA^0 will return 1 due to the 0 always returning 1 if anything is raised to it. But for_ `NA | TRUE` _where we are asking NA or True, we're just asking R to pick between "Idk" and "True" so it will always return_ `TRUE`. _Same logic applies to the_ `FALSE & NA`. _But_ `NA*0` _returns_ `NA` _because we still don't know if NA is a real number. If it turns out to be infinity, then R would return NaN but because we don't know if it's infinity it returns NA._

### 5.3.1 Exercises
**1. How could use `arrange()` to sort all missing values to the start?**
```{r}
arrange(flights, desc(is.na(dep_time)))
#desc means descending order. Because I nested is.na() in desc(), the output will have NA at the top of the list based on NAs in dep_time.
```


**2. Sort `flights` to find the most delayed flights. Find the flights that left the earliest.**
```{r}
arrange(flights, desc(dep_delay))
arrange(flights, dep_time)
#following the same logic as above, desc() will sort based on descending order of the departure delay.
```


**3. Sort `flights` to find the fastest flights.**
```{r}
arrange(flights, air_time)
```


**4. Which flights travelled the longest? Which travelled the shortest?**
```{r}
arrange(flights, distance)
```


### 5.4.1 Exercises
**1. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`.**
```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
```

```{r}
select(flights, starts_with("dep"), starts_with("arr"))
#Another way
```

**2. What happens if you include the name of a variable multiple times in a `select()` call?**
`select()` _just ignores the repeats._

**3. What does `one_of()` do? Why might it be helpful in conjunction with this vector?**
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
```

```{r}
select(flights, one_of(vars))
```


_Misnomer! This function just selects them all, not one of them as the name implies._

**4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?**
```{r}
select(flights, contains("TIME"))
```

_Surprising because usually R is case-sensitive. Can change that with_ `ignore.case`.


### 5.5.2 Exercises
**1. Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute because they're not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.**
_Convert the time to minutes since midnight using_ `mutate()` _or_ `transmutate()`.
```{r}
flight_minutes1 <- transmute(flights, minutes_dep_time = dep_time %/% 100 * 60 + dep_time %% 100, minutes_sched_dep_time = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100)

head(flight_minutes1)
```

**2. Compare `air_time` with `arr_time - dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?**
_Use_ `transmute()` _for a side-by-side comparison. I would expect to see that they are the same. Can use_ `ifelse()` _to handle some of the issues. The issues include handling time zones, taking off and landing on different days. Plotting to visualize the data really helps here._
```{r}
flight_minutes2 <- transmute(flights, minutes_dep_time = dep_time %/% 100 * 60 + dep_time %% 100, minutes_arr_time = arr_time %/% 100 * 60 + arr_time %% 100, minutes_air_time = air_time %/% 100 * 60 + air_time %% 100, calculated_time = ifelse(minutes_arr_time < minutes_dep_time, minutes_arr_time-minutes_dep_time + 1440, minutes_arr_time-minutes_dep_time), air_time)

head(flight_minutes2)
```

```{r}
ggplot(flight_minutes2, aes(x=air_time, y=calculated_time)) + geom_point(alpha=0.1)
```

_Better. I didn't address all problems that could have led to the difference._

**3. Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?**
`dep_delay` _should be the difference between_ `dep_time` _and_ `sched_dep_time`. _Be sure to convert to minutes and account for the issues in question 2 as well as those flights that leave early._

```{r}
flight_minutes3 <- transmute(flights, minutes_dep_delay = dep_delay %/% 100 * 60 + dep_time %% 100, minutes_sched_dep_time = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100, calculated_time = ifelse(minutes_dep_delay+120 < minutes_sched_dep_time, minutes_dep_delay-minutes_sched_dep_time + 1440, minutes_dep_delay-minutes_sched_dep_time))

head(flight_minutes3)
```

```{r}
ggplot(flight_minutes3, aes(x=minutes_dep_delay, y=minutes_dep_delay-minutes_sched_dep_time)) + geom_point(alpha=0.1)
```


**4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for `min_rank()`.**
```{r}
#make a smaller dataset for ease of use
ranking <- select(flights, carrier, flight, sched_dep_time, dep_delay, dep_time)

#do the ranking.
ranking <- mutate(ranking, delay_rank = min_rank(desc(dep_delay)))

#arrange based on the ranking
ranking <- arrange(ranking, delay_rank, carrier, flight)
head(ranking, n=10)
```

_The ranking isn't quite what I expected. I was expecting to have an actual ranking like 1, 2, 3, etc._

**5. What does `1:3 + 1:10` return? Why?**
```{r}
1:3 + 1:10
```

_It gives a weird error. And then also 1, 2, and 3 are added to part of_ `1:10` _because the objects are different lengths and_ `1:3` _can't be repeated through the_ `1:10` _because it's not a multiple. What about this:_

```{r}
1:3 + 1:9
```

_This works because 9 is a multiple of 3._

**6. What trigonometric functions does R provide?**
_Use_ `?trig` _to find out!_
```{r}
?Trig
```

