---
title: "Exploratory Data Analysis"
output: 
  html_document: 
    keep_md: yes
---

# Chapter 7: Exploratory Data Analysis

```{r, include=FALSE}
library(tidyverse)
library(diamonds)
library(nycflights13)
```

## 7.1 Introduction  
Use visualization and transformation to explore data in a systematic way = exploratory data analysis (EDA)  

## 7.2 Questions  
Goal of EDA is to develop understanding of data by using questions to guide investigation. Two types of questions will be useful:  
1. What type of variation occurs within my variables?  
2. What type of covariation occurs between my variables?  
Some vocab:  
* variable = qauntity, quality, or property that you can measure  
* value = state of a variable when it is measured; may change from measurement to measurement  
* observation = set of measurements made under similar conditions; contains several values associated with a different variable each. Also referred to as a data point  
* tabular data = set of values associated each with a variable and observation; tidy if each value is in own cell, each variable in own column, and each observation in own row  

## 7.3 Variation
* variation = tendency of values of a variable to change from measurement to measurement  
For example, if you measure a continuous vvariable twice, you will get two different results. Every variable has own pattern of variation.  

### 7.3.1 Visualising distributions
A variable is categorical if it can only take one of a small set of values. Usually saved as factors or character vectors. To examine these, use a bar chart:  
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

THe height of the bars displays how many observations occurred with each value on the x-axis. Compute these with `count()`:  
```{r}
diamonds %>% 
  count(cut)
```

A variable is continuous if it can take any of an infinite set of ordered values (think numbers and date-times). Examine with a histogram:  
```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

Compute:  
```{r}
diamonds %>% 
  count(cut_width(carat, 0.5))
```

Can set the width of the intervals with `binwidth` which is measured in units of the `x` variable:  
```{r}
smaller <- diamonds %>% 
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)
```

TO overlay multiple histograms use `geom_freqpoly()`:  
```{r}
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```
Now that data has been visualised, ask follow up questions like "what do you want to learn more about?" or "how could the data be misleading?".  

### 7.3.2 Typical Values
Useful questions:  
* Which values are most common?  
* Which values are rare? Does that match your expectations?  
* Can you see any unusual patterns?  

Example:  
```{r}
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

This plot leads to the questions:  
* Why are there more diamonds at whole carats and common fractions of carats?  
* Why are there more diamonds slightly to the right of each peak than there are slightly to the left?  
* Why are there no diamonds bigger than 3 carats?  

Clusters of similiar values suggest subgroups in data. Exploratory questions include:  
* How are the observations within each cluster similar to each other? How are they different?  
* How can you explain or describe the clusters?  
* Why might the appearance of clusters be misleading?  

Cluster example using Old Faithful:  
```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```

Clusters seem to be focused on short eruptions of about 2 minutes and long eruptions about 4-5 minutes.  

### 7.3.3 Unusual Values

OUtliers!  
Hard to see in a histogram:  
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

Only evidence is the unusually wide limits on the x-axis. So many observations in the common bins that the rare bins are so short you can't see them. Zoom in with `coord_cartesian()`:  
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```

`coord_cartesian()` also has `xlim()` for when you need to zoom in on the x-axis. Shows three unusual values. Select them:  
```{r}
unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)
unusual
```

Good practice to repeat analysis with and without outliers. If the outliers have minimal effect, they can be dropped. If they have a substantial effect, don't drop them without justification (ie figure out why they are there & disclose that in the write-up).

### 7.3.4 Exercises

**1. Explore the distribution of each of the x, y, and z variables in `diamonds`. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.**

**2. Explore the distribution of `price`. Do you discover anything unusual or surprising? (Hint: Carefully think about the `binwidth` and make sure you try a wide range of values.)**

**3. How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?**

**4. Compare and contrast `coord_cartesian()` vs `xlim()` or `ylim()` when zooming in on a histogram. What happens if you leave `binwidth` unset? What happens if you try and zoom so only half a bar shows?**

### 7.4.1 Exercises

**1. What happens to missing values in a historgram? What happens to missing values in a bar chart? Why is there a difference?**

**2. What does `na.rm = TRUE` do in `mean()` and `sum()`?**

### 7.5.1.1 Exercises
**1. Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.**

**2. What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?**

**3. Install the ggstance package, and create a horizontal boxplot. How does this compare to using `coord_flip()`?**

**4. One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using `geom_lv()` to display the distribution of price vs cut. What do you learn? How do you interpret the plots?**

**5. Compare and contrast `geom_violin()` with a facetted `geom_histogram()`, or a coloured `geom_freqpoly()`. What are the pros and cons of each method?**

**6. If you have a small dataset, it’s sometimes useful to use `geom_jitter()` to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to `geom_jitter()`. List them and briefly describe what each one does.**

### 7.5.2.1 Exercises

**1. How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut?**

**2. Use `geom_tile()` together with `dplyr` to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?**

**3. Why is it slightly better to use `aes(x = color, y = cut)` rather than `aes(x = cut, y = color)` in the example above?**

### 7.5.3.1 Exercises

**1. Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using `cut_width()` vs `cut_number()`? How does that impact a visualisation of the 2d distribution of carat and price?**

**2. Visualise the distribution of carat, partitioned by price.**

**3. How does the price distribution of very large diamonds compare to small diamonds. Is it as you expect, or does it surprise you?**

**4. Combine two of the techniques you’ve learned to visualise the combined distribution of cut, carat, and price.**

**5. Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of x and y values, which makes the points outliers even though their x and y values appear normal when examined separately.**
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```

**Why is a scatterplot a better display than a binned plot for this case?**
