---
title: "Chapter 5: Data Transformation, Continued"
output: 
  html_document: 
    keep_md: yes
---
```{r,include=FALSE}
library(tidyverse)
library(nycflights13)
```

##5.6 Grouped summaries with `summarise()`

`summarise()` collapses data frames to a single row. Example:
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

`summarise()` is more useful when used with `group_by()` because it analyzes individual groups instead of the whole dataset. Example:
```{r}
#Get average delay per date
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

###5.6.1 Combining multiple operations with the pipe

Can do it this way:
```{r}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")

# It looks like delays increase with distance up to ~750 miles 
# and then decrease. Maybe as flights get longer there's more 
# ability to make up delays in the air?
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
#> `geom_smooth()` using method = 'loess'
```

Where there are three steps to prepare the data: group flights by destination; summarise to compute distance, average delay, and number of flights; filter to remove noisy points and Honolulu airport, which is almost twice as far away as the next closest airport. Try again using the pipe (`%>%`):
```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```

This reads group, then summarise, then filter. This vastly improves the readability of the code. BUT `ggplot2` does not work with the pipe.

###5.6.2 Missing values

All aggregation functions have an `na.rm` argument. 

###5.6.3 Counts

It is good practice to either count `n()` or count non-missing values `sum(!is.na(x))` to make sure conclusions aren't drawn on small amounts of data. Example:
```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)
```

Get more insight if you plot the number of flights vs average delay:
```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

There is greater variation in the average delay when there are few flights. When you plot a mean (or other summary) vs group size, variation decreases as the sample size increases. Can also filter out groups that are small:
```{r}
delays %>% 
  filter(n > 25) %>% 
  ggplot(mapping = aes(x = n, y = delay)) + 
    geom_point(alpha = 1/10)
```

```{r, include=FALSE}
library(Lahman)

```

There's another common variation of this type of pattern. When plotting the skill of the batter (measured by batting average `ba`) against the number of opportunities to hit the ball (measured by at bat `ab`), there are two patterns:  
1. The variation in our aggregate decreases as we get more data points.  
2. There's a positive correlation between skill `ba` and opportunities to hit the ball `ab` because the teams control who gets to play.  
```{r}
# Convert to a tibble so it prints nicely
batting <- as_tibble(Lahman::Batting)

batters <- batting %>% 
  group_by(playerID) %>% 
  summarise(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )

batters %>% 
  filter(ab > 100) %>% 
  ggplot(mapping = aes(x = ab, y = ba)) +
    geom_point() + 
    geom_smooth(se = FALSE)
#> `geom_smooth()` using method = 'gam'
```


This has important implications for ranking. Sorting on `desc(ba)`, these players are lucky, not skilled:
```{r}
batters %>% 
  arrange(desc(ba))
```

###5.6.4 Useful summary functions

Measures of location: `mean(x)` which is the sum divided by the length or `median(x)` where 50% of `x` is above it and 50% is below it.  

Sometimes can be useful to combine aggregation with subsetting (which we haven't learned about yet):
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    avg_delay1 = mean(arr_delay),
    avg_delay2 = mean(arr_delay[arr_delay > 0]) # the average positive delay
  )
```

Measures of spread: `sd(x)` which is the standard deviation, `IQR(x)` which is the interquartile range, or `mad(x)` which is the median absolute deviation. `IQR(x)` and `mad(x)` are rough equivalents that are more useful if there are outliers in the data.

```{r}
# Why is distance to some destinations more variable than to others?
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(distance_sd = sd(distance)) %>% 
  arrange(desc(distance_sd))
```

Measures of rank: `min(x)`, `quantile(x, 0.25)` which is a generalization of the median (specifically here it will find a value of `x` that is greater than 25% of the values and less than the remaining 75%), or `max(x)`.
```{r}
# When do the first and last flights leave each day?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first = min(dep_time),
    last = max(dep_time)
  )
```

Measures of position: `first(x)`, `nth(x, 2)`, or `last(x)`. Work similarly to `x[1]`, `x[2]`, and `x[length(x)]` but lets you set a default value if that position does not exist ((i.e. you’re trying to get the 3rd element from a group that only has two elements).
```{r}
#Find first and last departure for each day
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first_dep = first(dep_time), 
    last_dep = last(dep_time)
  )
```

These are complemenary to filtering on ranks:
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  mutate(r = min_rank(desc(dep_time))) %>% 
  filter(r %in% range(r))
```

Counts: `n()` which takes no arguments and returns the size of the current group. For number of non-missing values use `sum(!is.na(x))`. For number of unique values use `n_distinct(x)`.
```{r}
# Which destinations have the most carriers?
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers = n_distinct(carrier)) %>% 
  arrange(desc(carriers))
```

If all you want is count:
```{r}
not_cancelled %>% 
  count(dest)
```

If you want to weight a variable:
```{r}
#Total number of miles a plane flew
not_cancelled %>% 
  count(tailnum, wt = distance)
```

Counts and proportions of logical values: `sum(x>10)` or `mean(y==0)`. When used with numeric functions `TRUE` is converted to 1 and `FALSE` to 0 making `sum(x)` give number of `TRUE` in `x` and `mean(x)` gives proportion.
```{r}
# How many flights left before 5am? (these usually indicate delayed
# flights from the previous day)
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(n_early = sum(dep_time < 500))

# What proportion of flights are delayed by more than an hour?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(hour_perc = mean(arr_delay > 60))
```

###5.6.5 Grouping by multiple variables

When grouping by multiple variables, each summary peels off one level of the grouping, making it easy to roll up a data set:
```{r}
daily <- group_by(flights, year, month, day)
(per_day   <- summarise(daily, flights = n()))

(per_month <- summarise(per_day, flights = sum(flights)))

(per_year  <- summarise(per_month, flights = sum(flights)))
```

It's ok to progressively roll up sums and counts, but need to think about weighting means and variances and not possible to do it exactly for rank-based statistics like the median. The sum of groupwise sums is the overall sum, but the median of groupwise medians is not the overall median.

###5.6.6 Ungrouping

To  remove grouping, and return to operations on ungrouped data, use `ungroup()`.
```{r}
daily %>% 
  ungroup() %>%             # no longer grouped by date
  summarise(flights = n())  # all flights
```

###5.6.7 Exercises

**1. Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios:**  

* A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.
```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
#First way
not_cancelled %>% group_by(year, month, day) %>% summarise(median = median(filter(arr_delay==15 | arr_delay==-15)))

#Second way
not_cancelled %>% group_by(year, month, day) %>% summarise(median = median(filter(dep_delay==15 | dep_delay==-15)))

#Third way
not_cancelled %>% group_by(year, month, day) %>% filter(arr_delay==15, arr_delay==-15) %>% summarise(arr_delay)

#Fourth way
not_cancelled %>% group_by(year, month, day) %>% filter(dep_delay==15, dep_delay==-15) %>% summarise(dep_delay)
```

* A flight is always 10 minutes late. 
```{r}
#First way
not_cancelled %>% group_by(year, month, day) %>% filter(arr_delay==10) %>% summarise(arr_delay)

#Second way
not_cancelled %>% group_by(year, month, day) %>% filter(dep_delay==10) %>% summarise(dep_delay)

```

* A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.  
```{r}
#First way
not_cancelled %>% group_by(year, month, day) %>% summarise(median = median(filter(arr_delay==30 | arr_delay==-30)))

#Second way
not_cancelled %>% group_by(year, month, day) %>% summarise(median = median(filter(dep_delay==30 | dep_delay==-30)))

#Third way
not_cancelled %>% group_by(year, month, day) %>% filter(arr_delay==30, arr_delay==-30) %>% summarise(arr_delay)

#Fourth way
not_cancelled %>% group_by(year, month, day) %>% filter(dep_delay==30, dep_delay==-30) %>% summarise(dep_delay)
```

* 99% of the time a flight is on time. 1% of the time it's 2 hours late. 
```{r}
#First way
not_cancelled %>% group_by(year, month, day) %>% filter(arr_delay >= 120) %>% summarise(onepercent = quantile(arr_delay, 0.01))

#Second way
not_cancelled %>% group_by(year, month, day) %>% filter(dep_delay >= 120) %>% summarise(onepercent = quantile(dep_delay, 0.01))

```


**Which is more important, arrival delay or departure delay?**  
_Personally I think arrival delay is more important because it is mor inconvenient than bein delayed departing._

**2. Come up with another approach that will give you the same outpt as `not_cancelled %>% count(tailnum, wt = distance)` without using `count()`.**

**3. Our definition of cancelled flights `is.na(dep_delay) | is.na(arr_delay)` is slightly suboptimal. Why? Which is the most important column?**  

**4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?**

**5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights %>% group_by(carrier, dest) %>% summarise(n())`.**

**6. What does the `sort` argument to `count()` do. When might you use it?**

##5.7 Grouped mutates and filters  

###5.7.1 Exercises  

**1. Refer back to the lists of useful mutate and filtering functions. Describe how each operation changes when you combine it with grouping.**  

**2. Which plane (`tailnum`) has the worst on-time record?**  

**3. What time of day should you fly if you want to avoid delays as much as possible?**

**4. For each destination, compute the total minutes of delay. For each, flight, compute the proportion of the total delay for its destination.**

**5. Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using `lag()` explore how the delay of a flight is related to the delay of the immediately preceding flight.**  

**6. Look at each destination. Can you find flights that are suspiciously fast? (i.e. flights that represent a potential data entry error). Compute the air time a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?**  

**7. Find all destinations that are flown by at least two carriers. Use that information to rank the carriers.**  

**8. For each plane, count the number of flights before the first delay of greater than 1 hour.**  

#Chapter 6: Workflow: scripts  

###6.3 Practice  

**1. Go to the RStudio Tips twitter account, https://twitter.com/rstudiotips and find one tip that looks interesting. Practice using it!**

**2. What other common mistakes will RStudio diagnostics report? Read https://support.rstudio.com/hc/en-us/articles/205753617-Code-Diagnostics to find out.**
