---
title: "R for Data Science, Chapter 3"
output: 
  html_document: 
    keep_md: yes
---

# Data Visualization 

We will be using the _tidyverse_ dataset:
```{r include = FALSE}
library(tidyverse)
```

### 3.2.4 Exercises 
**1. Run ggplot(data = mpg) what do you see?**  
**2. How many rows are in mtcars? How many columns?**   
**3. What does the drv variable describe? Read the help for ?mpg to find out.**  
**4. Make a scatterplot of hwy vs cyl.**  
**5. What happens if you make a scatterplot of class vs drv. Why is the plot not useful?**  

### 3.3.1 Exercises 
**1. What's wrong with this code? Why are the points not blue?**  
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

**2. Which variables in mpg are categorical? Which variables are continuous? Hint: type ?mpg to read the documentation for the dataset. How can you see this information when you run mpg?**  
**3. Map a continous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs continous variables?**  
**4. WHat happens if you map the same variable to multiple aesthetics?**  
**5. What does the stroke aesthetic do? What shapes does it work with? Hint: use ?geom_point.**  
**6. What happens if you map an aesthetic to something other than a variable name, like aes(color = displ < 5)?**  

### 3.5.1 Exercises  
**1. What happens if you facet on a continuous variable?**  
**2. What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?**  
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
```

**3. What plots does the following code make? What does . do?**  
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ .)
```

**4. Take the first faceted plot in this section:**
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
**What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?**  
**5. Read ?facet_wrap. What does nrow do? What does ncol do? WHat other options control the layout of the individual panels? Why doesn't facet_grid() have nrow and ncol variables?**  
**6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?**  